### 2023-07-12 intro.tex

This localization decision affects how the system recovers as it proceeds into the let body. 
If localization had assumed that the \li{then} branch were correct, as for example in \li{merlin}, then 
\li{x : Bool} and an error should be reported on only the second use of \li{x}. 
If the \li{else} branch were chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
 In either case, this error might mislead the programmer if the earlier localization guess was incorrect.
If the inconsistency were localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error at either use of \li{x}, 
so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing.
In any of these cases, we would like to be able to recover and 
localize the type inconsistency on the string addend because the \li{+} operator is integer addition in Hazel.
Recovering from this error, we can assume that both branch types in the second conditional will have \li{Int} type, so no 
error is marked on the conditional as a whole.

@@

Section~\ref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays to the user.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e. its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g. with destructuring patterns and System F-style polymorphism, 
that serve as case studies of the more general approach that we hope will be adopted by language designers defining practical type systems.

\cref{sec:calculus-hazel} describes our own effort in this direction, which is to scale up the marked lambda calculus as the new basis for Hazel, a typed functional programming environment that fully solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed. Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as in Agda \cite{norell:thesis}, GHC Haskell \cite{haskell-holes}, Idris \cite{DBLP:journals/jfp/Brady13}, and others), syntax error recovery \cite{medeiros20,sorkin11}, and structure editing \cite{HazelnutPOPL,teitelbaum_cornell_1981}. 
Hazel has both a textual syntax and a structure editor \cite{HazelnutPOPL,DBLP:conf/vl/Moon023}. 
%offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e. a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}. 
As a secondary contribution, 
\cref{sec:calculus-hazelnut} describes how (1)  total marking can resolve the problem of undefined behavior in the Hazelnut structure editor calculus, 
and (2) integrating marking with typed structure editing allows us to incrementally re-mark only where necessary based on the edit location and the type and binding structure.

>>>>

This localization decision affects how the system recovers as it proceeds into the let body. 
If localization had assumed that the \li{then} branch were correct, as for example in \li{merlin}, then 
\li{x : Bool} and an error should be reported on only the second use of \li{x}. 
If the \li{else} branch were chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
 In either case, this error might mislead the programmer if the earlier localization guess was incorrect.
If the inconsistency were localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
\li{x} does not have a known type,
though it \emph{is} known to be bound. 
There is no definitive type or binding error at either use of \li{x}, 
so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing \cite{Siek06a}.
In any of these cases, we would like to be able to recover and 
localize the type inconsistency on the string addend because the \li{+} operator is integer addition in Hazel.
Recovering from this error, we can assume that both branch types in the second conditional will have \li{Int} type, so no 
error is marked on the conditional as a whole.

@@

\Cref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e. its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g. with destructuring patterns and System F-style polymorphism, 
that serve as case studies of the general approach that we hope this paper will be adopted by language designers defining practical type systems.

\Cref{sec:calculus-hazel} describes our own effort in this direction, which is to scale up the marked lambda calculus as the new basis for Hazel, a typed functional programming environment that fully solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed. Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as in Agda \cite{norell:thesis}, GHC Haskell \cite{haskell-holes}, Idris \cite{DBLP:journals/jfp/Brady13}, and others), syntax error recovery \cite{medeiros20,sorkin11}, and structure editing \cite{HazelnutPOPL,teitelbaum_cornell_1981}. 
Hazel has both a textual syntax and a structure editor \cite{HazelnutPOPL,DBLP:conf/vl/Moon023}. 
%offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e. a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}. 
As a secondary contribution, 
\cref{sec:calculus-hazelnut} describes how (1)  total marking can resolve the problem of undefined behavior in the Hazelnut structure editor calculus, 
and (2) integrating marking with typed structure editing allows us to incrementally re-mark only where necessary based on the edit location and the type and binding structure.
